An Algorithm for Searching Evolving Graphs in Linear Algebraic Language

### 2016-01-23 paper.tex
Clean up presentation of linear algebra

@@

This algebraic representation allows us to exploit a graphical interpretation of
matrix--vector products involving the adjacency matrix~\cite{kegi11}.
If $A$ is the adjacency matrix of a (static) graph $G$ and $e_k$ is $k$th
elementary unit vector, then the nonzero entries of $A^T e_k$ have indices that
are neighbors of $k$.
The algebraic formulation of BFS on evolving graphs follows similarly, but
requires a new kind of matrix--vector product, $\odot$, defined by
\[
A^T \!\odot b =
\begin{cases}
b & \mbox{if $A^Tb \ne 0$,} \\
0 & \mbox{otherwise.}
\end{cases}
\]
The forward neighbors of a temporal node $(k, t_1)$ in $A_n$ are then be determined
from the indices and time stamps of the nonzero elements in the sequence
\begin{equation}
\label{eq:out_ne}
\big\langle (A^{[1]})^Te_k, \; (A^{[2]})^T\!\odot e_k, \; \ldots \; (A^{[n]})^T\!\odot e_k\big\rangle.
\end{equation}
The nonzero entries of the first vector represent forward neighbors that are on
the same time stamp $t_1$, whereas all other nonzero entries represent forward
neighbors that are advanced in time but remain on the same node $k$.
The quantity \eqref{eq:out_ne} therefore encodes a BFS tree of depth 2,
as its nonzero entries are labeled by all temporal nodes of distance 1 from
$(k, t_1)$.

@@

Let $(A^{[i]})^Tb = c_i$.
At the second step, we need to compute the following components
\begin{align}
  \label{eq:5}
 \big\langle (A^{[1]})^Tc_1,   (A^{[2]})^T\odot c_1, \ldots,  (A^{[n]})^T\odot c_1  \big\rangle  \\
\big\langle (A^{[2]})^Tc_{2}, \ldots, (A^{[n]})^T\odot c_{2} \big\rangle \\
 \ldots  \\
  \big\langle (A^{[n]})^Tc_n \big\rangle
\label{eq:6}
\end{align}
If we sum these resulting vectors according to time stamps, we obtain
the forward-neighbors of the nodes computed at step $1$.
For $n =2$, we have
\[
\Big\langle
(A^{[1]})^Tc_1,~
(A^{[2]})^T\odot c_1 + (A^{[2]})^T c_2
\Big\rangle.
\]
If we form a block triangular matrix $\bm{A_n}^T$
\[
\bm{A_n}^T =
\begin{bmatrix}
(A^{[1]})^T &   0            & \cdots  &  & 0 \\
\sigma(A^{[2]})^T & (A^{[2]})^T & 0         &\cdots  & \vdots \\
\sigma(A^{[3]})^T & \sigma(A^{[3]})^T & (A^{[3]})^T & \cdots \\
   \vdots              &                              &      \vdots            &  &  \vdots \\
\sigma(A^{[n]})^T &                  &       \cdots                      &  & (A^{[n]})^T
\end{bmatrix},
\]
where  $\sigma(A^{[i]})^Tb = (A^{[i]})^T\odot b$, and a block vector $\bm{b}$
\[
\bm{b}^T = [b^T, 0, \cdots, 0  ],
\]
then \eqref{eq:out_ne} is equivalent to $\bm{A_n}^T\bm{b}$ and
summing \eqref{eq:5}-\eqref{eq:6} is equivalent to $\bm{A_n}^T(\bm{A_n}^T\bm{b})$.
In general, we can define an evolving-graph BFS using
block vector-matrix multiplications. The algorithm is presented in Algorithm
\ref{alg:bfsla}.

\begin{algorithm}[h]
 \SetKwFor{For}{for}{}{end}
 \SetKwFor{While}{while}{}{end}
 \SetKwFor{If}{if}{}{end}
 \SetKwFunction{nei}{forward-neighbors}
 \SetKwFunction{enqueue}{enqueue}
 \SetKwFunction{nz}{nonzeros}
 \SetKwFunction{intersect}{intersect}
 \SetKwFunction{map}{map}
 \SetKwProg{Fn}{function}{}{end}
 \Fn{ABFS($A_n, (s,t)$)}{
   Form $\bm{A}^T$ from $A_n$.\\
   Let $m$ be the dimension of  $A^{[1]}$. \\
   $b$ is a sparse vector of length $mn$ \\
   $b(s+mt) =1$ \\
   $i = 1$ \\
   $dict.(s,t) = 0$ \\
   \While{\nz{$b$} $\ne \emptyset$}{
       $\bm{b} = \bm{A}^T\bm{b}$ \\
       \For{$i$ in \nz{$b$}}{
         $node =$ \map{$i$} \\
         \If{$node$ in $dict$}{
           $b(i) = 0$ \\
         }
      }
      $nodes =$\map{$b$}\\
       \For{$node$ in $nodes$}{
           $dict.node = i$
         }

       $i = i + 1$
       }
  \textbf{return} $dict$ \\
}
 \caption{Algebraic evolving-graph BFS.
Let $A_n$ be the corresponding adjacency matrices of $G_n$ and
$(s,t)$ a node of $G_n$. As in Algorithm \ref{alg:bfs}, we use a dictionary $dict.(v_i, t_i)$
to store $(v_i, t_i)$'s distance from $(s,t)$. Function \texttt{nonzeros(v)}
returns the nonzero indices of a given vector  $v$ and function \texttt{map(b)}
maps a block vector's indices to their corresponding nodes. The algorithm  returns
$dict$ which stores all the reachable active nodes and their distance from $(s,t)$.}
 \label{alg:bfsla}
\end{algorithm}

Algorithm \ref{alg:bfsla} outlines the basic idea of an algebraic evolving-graph BFS.
But in practice we don't need to form $\bm{A}$ explicitly at step $2$. By realizing the dependency between these sparse matrices,
we can compute  the result in parallel.

@@

\IEEEproof
We shall check the two algorithms are equivalent in each module of
the method.
The initialization steps are trivially equivalent.
By the construction of $\bm{A^T}$,
step $9$ of Algorithm \ref{alg:bfsla} computes the forward-neighbors
of all the frontier nodes, which is equivalent to step $8$-$9$
of Algorithm \ref{alg:bfs}. Step $10$-$15$ of Algorithm \ref{alg:bfsla} ensures
active nodes will not be visited twice, which is equivalent to step $10$
of Algorithm \ref{alg:bfs}. Finally, a dictionary $dict$ is used by both algorithms
to keep record of the distance of each node from the source $(s,t)$.
\endproof

>>>>

\subsection{Evolving graphs as sequences of adjacency matrices}

@@

This algebraic representation allows us to exploit a graphical interpretation of
matrix--vector products involving the adjacency matrix~\cite{kegi11}.
If $A$ is the adjacency matrix of a (static) graph $G$ and $e_k$ is the $k$th
elementary unit vector, then the nonzero entries of $A^T e_k$ have indices that
are neighbors of $k$.
The algebraic formulation of BFS on evolving graphs follows similarly, but
requires a new kind of matrix--vector product, $\odot$, defined by
\[
A^T \!\odot b =
\begin{cases}
b & \mbox{if $A^Tb \ne 0$,} \\
0 & \mbox{otherwise.}
\end{cases}
\]
The forward neighbors of a temporal node $(k, t_1)$ in $A_n$ can then be determined
from the indices and time stamps of the nonzero elements in the sequence
\begin{equation}
\label{eq:out_ne}
\big\langle (A^{[1]})^Te_k, \; (A^{[2]})^T\!\odot e_k, \; \ldots \; (A^{[n]})^T\!\odot e_k\big\rangle.
\end{equation}
The nonzero entries of the first vector represent forward neighbors that are on
the same time stamp $t_1$, whereas all other nonzero entries represent forward
neighbors that are advanced in time but remain on the same node $k$.
The quantity \eqref{eq:out_ne} therefore encodes a BFS tree of depth 2,
as its nonzero entries are labeled by all temporal nodes of distance 1 from
$(k, t_1)$.

@@

\subsection{Evolving graphs as a blocked adjacency matrix}

The proof of Theorem~\ref{thm:correct} provided a construction for representing
an evolving graph $G_n$ by a static graph $G$ with nodes corresponding to active
nodes of $G_n$. The adjacency matrix $\mathbf A$ of $G$ then has block structure
which is useful for understanding the nature of the $\odot$ operation.

Consider two iterations of the BFS on $G_n$ with root $(k, t_1)$. The second
iteration requires computing the sequences

\begin{subequations}
\begin{align}
  \label{eq:5}
 \big\langle (A^{[1]})^Tc_1,   (A^{[2]})^T\odot c_1, \ldots,  (A^{[n]})^T\odot c_1  \big\rangle  \\
\big\langle (A^{[2]})^Tc_{2}, \ldots, (A^{[n]})^T\odot c_{2} \big\rangle \\
 \ldots  \\
  \big\langle (A^{[n]})^Tc_n \big\rangle
\label{eq:6}
\end{align}
\end{subequations}
%
where $c_i = (A^{[i]})^Te_k$. Summing resultant vectors that share the same
time stamp, we obtain vectors whose nonzero elements have indexes labeled by
the forward neighbors of the nodes computed at step $1$.

Define the block triangular adjacency matrix
\[
\bm{A_n}^T =
\begin{bmatrix}
(A^{[1]})^T &   0            & \cdots  &  & 0 \\
\sigma(A^{[2]})^T & (A^{[2]})^T & 0         &\cdots  & \vdots \\
\sigma(A^{[3]})^T & \sigma(A^{[3]})^T & (A^{[3]})^T & \cdots \\
   \vdots              &                              &      \vdots            &  &  \vdots \\
\sigma(A^{[n]})^T &                  &       \cdots                      &  & (A^{[n]})^T
\end{bmatrix},
\]
where $\sigma(A^{[i]})^Tb = (A^{[i]})^T\odot b$, and the block vector
\[
\bm{b}^T = [b^T, 0, \cdots, 0  ].
\]
Then, \eqref{eq:out_ne} is equivalent to the matrix--vector product
$\bm{A_n}^T\bm{b}$ and the summation over \eqref{eq:5}-\eqref{eq:6} is equivalent
to $\bm{A_n}^T(\bm{A_n}^T\bm{b})$.

\subsection{The algebraic formulation of BFS on evolving graphs}

The blocked matrix--vector products introduced in the previous section allows us
to write down an elegant algebraic formulation of BFS on evolving graphs, as
presented in Algorithm~\ref{alg:bfsla}. Note that $\bm A^T$ need never be formed
explicitly, as only its matrix--vector product is required in the algorithm.

\begin{algorithm}[h]
 \SetKwFor{For}{for}{}{end}
 \SetKwFor{While}{while}{}{end}
 \SetKwFor{If}{if}{}{end}
 \SetKwFunction{nei}{forward-neighbors}
 \SetKwFunction{enqueue}{enqueue}
 \SetKwFunction{nz}{nonzeros}
 \SetKwFunction{intersect}{intersect}
 \SetKwFunction{map}{map}
 \SetKwProg{Fn}{function}{}{end}
 \Fn{ABFS($A_n, (v_1,t_1)$)}{
   Form $\bm{A}^T$ from $A_n$.\\
   $\bm b = e_{v_1}$ \\
   $k = 1$ \\
   $reached[(v_1,t_1)] = 0$ \\
   \While{\nz{$b$} $\ne \emptyset$}{
       $\bm{b} = \bm{A}^T\bm{b}$ \\
       \For{$i \in$ \nz{$b$}}{
         \If{\map{$k$} $\in reached$}{
           $b_k = 0$ \\
         }
      }
      $nodes =$\map{$b$}\\
       \For{$node \in reached$}{
           $reached[node] = k$ \\
         }
       $k = k + 1$
       }
  \textbf{return} $reached$ \\
}
\caption{An algebraic formulation of BFS on evolving graphs.
Given $A_n$, the adjacency matrix representation of $G_n$ and $(v_1,t_1)$, a
node of $G_n$, returns $reached$ as defined in Algorithm~\ref{alg:bfs}.
The function \texttt{nonzeros(v)} returns the nonzero indices of the vector $v$,
and the function \texttt{map(b)} maps a block vector's indices to their corresponding
nodes.}
\label{alg:bfsla}
\end{algorithm}

@@

\IEEEproof
The initialization steps are trivially equivalent.
At the beginning of iteration $k$, the block vector $\bm b$ represents the
frontier nodes encoding the $frontier$ set of Algorithm~\ref{alg:bfs}.
The matrix--vector product $\bm A^T \bm b$ encodes the forward neighbors of all
the frontier nodes. Subequently, active nodes that have already been visited in
previous iterations are zeroed out of the new $\bm b$.
\endproof
